---
title: Polling in the 2022 Forecast
author: Ethan Kelly
date: '2022-09-26'
slug: polling-forecasts
categories: []
tags: []
authors: []
description: ''
externalLink: ''
series: []
---


```{r, echo = FALSE, message = FALSE, include = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")
```

```{r, include = FALSE}
# Load Libraries
library(tidyverse)
library(usmap)
library(sf)
library(rmapshaper)
library(car)
library(moderndive)
```

```{r, include = FALSE}
# load datasets for extension 1
general_polls_1942_2020 <- read_csv("GenericPolls1942_2020.csv")
generic_ballot_averages_2018_2022 <- read_csv("538_generic_ballot_averages_2018-2022.csv")


# for model
house_election_data <- read_csv('house_popvote_seats.csv') 
RDI_monthly <- read_csv('RDI_monthly.csv') 
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# filter out variables from election data
house_pres_data <- house_election_data %>%
  select(year, R_seats, D_seats, Other_seats, winner_party, R_majorvote_pct, D_majorvote_pct, president_party, H_incumbent_party, H_incumbent_party_winner, H_incumbent_party_majorvote_pct)

# filter the RDI data
rdi_edit <- RDI_monthly %>%
  mutate(year = substr(date, 1, 4),
         month = substr(date, 6, 7),
         year = as.numeric(year),
         month = as.numeric(month)) %>%
  rename(rdi = DSPIC96)

# further edit the rdi dataset
rdi_change <- rdi_edit %>%
  filter(month %in% c(4,11)) %>%
  group_by(year) %>%
  mutate(rdi_change = (rdi - lag(rdi)),
         rdi_change_pct = (rdi/lag(rdi) - 1) * 100,
         midterm_year = if_else(
           year %in% c(2018,2014,2010,2006,2002,1998,1994,1990,1986,1982,1978,1974,1970,1966,1962,1958,1954,1950), "midterm", "pres")) %>%
  select(year, rdi, rdi_change, rdi_change_pct, midterm_year) %>%
  na.omit()

# join with house election data
rdi_election_data_dem_joined <- inner_join(house_pres_data, rdi_change, by = c("year" = "year"))

# filter for only dem house
rdi_election_data_dem_control_house <- rdi_election_data_dem_joined %>%
  filter(H_incumbent_party == "D")
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# Look into Dem vote share for Dem house incumbent
# correlation coefficient
cor(rdi_election_data_dem_control_house$rdi_change_pct,
    rdi_election_data_dem_control_house$D_majorvote_pct)

# create linear model for RDI House Incumbent under Dem House data
rdi_Hincumbent_votes_Dhouse_lm <- lm(D_majorvote_pct ~ rdi_change_pct, 
                              data = rdi_election_data_dem_control_house)

summary(rdi_Hincumbent_votes_Dhouse_lm)

# According to BEA, cpi change from 4/1/22 to 8/1/22 is a 1.91% increase
rdi_2022 <- data.frame(rdi_change_pct = 1.91)

# predict House incumbent vote share with current RDI 
predict(rdi_Hincumbent_votes_Dhouse_lm,
        rdi_2022,
        interval = "prediction",
        level = 0.95)
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# visualize rdi data for Dems vote share under Dem Pres
ggplot(data = rdi_election_data_dem_control_house,
       mapping = aes(x = rdi_change_pct,
                     y = D_majorvote_pct,
                     label = year)) +
  geom_text() +
  geom_smooth(method = "lm", se = TRUE) +
  geom_hline(yintercept = 50, lty = 2) +
  geom_vline(xintercept = 0.00, lty = 2) +
  labs(x = "RDI Growth From April to November in Midterm Years Under a Democratic House",
       y = "Democratic Party Vote Share",
       subtitle = "p = 0.0025, Adjusted R-squared = 0.5775",
       title = "Y = 55.3872 - 2.49x") +
  theme_bw() +
  theme(axis.text = element_text(size = 10))
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# Look into Dem seat share for Dem House incumbent
# correlation coefficient
cor(rdi_election_data_dem_control_house$rdi_change_pct,
    rdi_election_data_dem_control_house$D_seats)

# create linear model for RDI House Incumbent under Dem House data
rdi_Hincumbent_seats_Dhouse_lm <- lm(D_seats ~ rdi_change_pct, 
                              data = rdi_election_data_dem_control_house)

summary(rdi_Hincumbent_seats_Dhouse_lm)

# Create data frame of current statistics

# predict House incumbent seat share with current RDI 
predict(rdi_Hincumbent_seats_Dhouse_lm,
        rdi_2022,
        interval = "prediction",
        level = 0.95)
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# visualize rdi data for Dems seat share under Dem House
ggplot(data = rdi_election_data_dem_control_house,
       mapping = aes(x = rdi_change_pct,
                     y = D_seats,
                     label = year)) +
  geom_text() +
  geom_smooth(method = "lm", se = TRUE) +
  geom_hline(yintercept = 218, lty = 2) +
  geom_vline(xintercept = 0.00, lty = 2) +
  labs(x = "RDI Growth From April to November in Midterm Years Under a Democratic House",
       y = "Democratic Party Seat Share",
       subtitle = "p = 0.0025, Adjusted R-squared = 0.4957",
       title = "Y = 269.588 - 20.22x") +
  theme_bw() +
  theme(axis.text = element_text(size = 10))
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# predict House incumbent vote share with current RDI 
predict(rdi_Hincumbent_votes_Dhouse_lm,
        rdi_2022,
        interval = "prediction",
        level = 0.95)
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# predict House incumbent seat share with current RDI 
predict(rdi_Hincumbent_seats_Dhouse_lm,
        rdi_2022,
        interval = "prediction",
        level = 0.95)
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# True = line
# prediction = dots
# visualize rdi data for Dems vote share under Dem House
ggplot(data = rdi_election_data_dem_control_house,
       mapping = aes(x = year,
                     y = D_majorvote_pct)) +
  geom_line(aes(y = D_majorvote_pct,),
            size = 1.5) +
  geom_point(y = predict(rdi_Hincumbent_votes_Dhouse_lm, rdi_election_data_dem_control_house),
             color = "red",
             size = 2) +
  geom_hline(yintercept = 50, lty = 2) +
  labs(x = "Year",
       y = "Democratic House Incumbent Party Vote Share ",
       subtitle = "Vote Share Based on Real Disposable Income Change",
       title = "True Results (Line) vs Model Prediction (Dots)") +
  theme_bw() +
  theme(axis.text = element_text(size = 10))


# visualize rdi data for Dems seat share under Dem House
ggplot(data = rdi_election_data_dem_control_house,
       mapping = aes(x = year,
                     y = D_seats)) +
  geom_line(aes(y = D_seats,),
            size = 1.5) +
  geom_point(y = predict(rdi_Hincumbent_seats_Dhouse_lm, rdi_election_data_dem_control_house),
             color = "red",
             size = 2) +
  geom_hline(yintercept = 50, lty = 2) +
  labs(x = "Year",
       y = "Democratic House Incumbent Party Vote Share ",
       subtitle = "Seat Share Based on Real Disposable Income Change",
       title = "True Results (Line) vs Model Prediction (Dots)") +
  theme_bw() +
  theme(axis.text = element_text(size = 10))
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# Residuals for Vote Share
ggplot(data = rdi_Hincumbent_votes_Dhouse_lm,
       mapping = aes(x = rdi_Hincumbent_votes_Dhouse_lm$model$D_majorvote_pct -
rdi_Hincumbent_votes_Dhouse_lm$fitted.values)) +
  geom_histogram(bins = 15) +
  labs(x = "Residuals Between True and Predicted Vote Shares",
       y = "Frequency",
       subtitle = "Vote Share Based on Real Disposable Income Change",
       title = "Error in Model") +
  theme_bw() +
  theme(axis.text = element_text(size = 10))

# Residuals for Seat Share
ggplot(data = rdi_Hincumbent_seats_Dhouse_lm,
       mapping = aes(x = rdi_Hincumbent_seats_Dhouse_lm$model$D_seats -
rdi_Hincumbent_seats_Dhouse_lm$fitted.values)) +
  geom_histogram(bins = 15) +
  labs(x = "Residuals Between True and Predicted Seat Shares",
       y = "Frequency",
       subtitle = "Seat Share Based on Real Disposable Income Change",
       title = "Error in Model") +
  theme_bw() +
  theme(axis.text = element_text(size = 10))
```

## Polling and Election Forecasting in Practice

Below, I explore the relationship between the Democrats' Polling Margin from an aggregate of polls conducted within 50 days of an election and the Democrats' Actual Vote/Seat Share in the election. 

As evidenced below from the strong correlation and Adjusted R-Squared that explains how well the model explains variability in the data, the Democrats' Polling Margin from an aggregate of polls conducted within 50 days of an election is a strong predictor of their vote/seat share outcome. Including the RDI change from April to November of each election year does not seem to play a huge factor in improving this correlation. However, it does not hurt the modeling.

#### Vote Share
```{r}
# vote share

# join rdi and pres with general ballot
polls_rdi_election <- left_join(general_polls_1942_2020, rdi_election_data_dem_joined, by = c("year" = "year"))

# filter down
polls_rdi_election_vote_filter <- polls_rdi_election %>%
  mutate(poll_margin = dem - rep,
         D_votemargin = D_majorvote_pct - R_majorvote_pct,
         poll_vote_diff = dem - D_majorvote_pct) %>%
  drop_na(rdi_change_pct) %>%
  filter(days_until_election > 0,
         days_until_election < 50,
         year != 2008)

# create averages by election year to avoid bias in polls
polls_rdi_election_vote_sum <- polls_rdi_election_vote_filter %>%
  group_by(year) %>%
  summarize(mean_dem = mean(dem),
         mean_rep = mean(rep),
         mean_dem_margin = mean(poll_margin),
         mean_days_until_election = mean(days_until_election),
         rdi_change_pct = mean(rdi_change_pct),
         D_majorvote_pct = mean(D_majorvote_pct),
         D_votemargin = mean(D_votemargin),
         poll_vote_diff = mean(poll_vote_diff))
  
  
# plot graph
ggplot(data = polls_rdi_election_vote_sum,
       mapping = aes(x = mean_dem_margin,
                     y = D_majorvote_pct,
                     color = rdi_change_pct,
                     label = year)) +
  geom_text(size = 3) + 
  geom_hline(yintercept = 50, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_smooth(method = lm) +
  labs(x = "Mean Polling Margin of Democratic Party",
       y = "Actual Vote Share of Democratic Party",
       title = "p = 5e-09, Adjusted R-squared = 0.7393")
```

```{r, include = FALSE}
# correlation coefficient
cor(polls_rdi_election_vote_sum$D_majorvote_pct,
    polls_rdi_election_vote_sum$mean_dem_margin)

# create linear model
rdi_polls_sum_vote_lm <- lm(D_majorvote_pct ~ mean_dem_margin + rdi_change_pct,
                            data = polls_rdi_election_vote_sum)

summary(rdi_polls_sum_vote_lm)

# Create data frame of current statistics
# dem margin: RCP average = D+0.3
# According to BEA, cpi change from 4/1/22 to 8/1/22 is a 1.91% increase
poll_and_rdi_data_vote <- data.frame(mean_dem_margin = 0.3,
                                rdi_change_pct = 1.91)

# predict Dem vote share with current RDI and Dem Margin
predict(rdi_polls_sum_vote_lm,
        poll_and_rdi_data_vote,
        interval = "prediction",
        level = 0.95)
```

#### Seat Share
```{r}
# seat share

# join rdi and pres with general ballot
polls_rdi_election <- left_join(general_polls_1942_2020, rdi_election_data_dem_joined, by = c("year" = "year"))

# filter down
polls_rdi_election_seat_filter <- polls_rdi_election %>%
  mutate(poll_margin = dem - rep,
         D_seat_margin = D_seats - R_seats) %>%
  drop_na(rdi_change_pct) %>%
  filter(days_until_election > 0,
         days_until_election < 50,
         year != 2008)

# create averages by election year to avoid bias in polls
polls_rdi_election_seat_sum <- polls_rdi_election_seat_filter %>%
  group_by(year) %>%
  summarize(mean_dem = mean(dem),
         mean_rep = mean(rep),
         mean_dem_margin = mean(poll_margin),
         mean_days_until_election = mean(days_until_election),
         rdi_change_pct = mean(rdi_change_pct),
         D_seats = mean(D_seats))
  
  
# plot graph
ggplot(data = polls_rdi_election_seat_sum,
       mapping = aes(x = mean_dem_margin,
                     y = D_seats,
                     color = rdi_change_pct,
                     label = year)) +
  geom_text(size = 3) + 
  geom_smooth(method = "lm") +
  geom_hline(yintercept = 218, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  labs(x = "Mean Polling Margin of Democratic Party",
       y = "Actual Seat Share of Democratic Party",
       title = "p = 2.778e-09, Adjusted R-squared = 0.7504")
```

```{r, include = FALSE}
# correlation coefficient
cor(polls_rdi_election_seat_sum$D_seats,
    polls_rdi_election_seat_sum$mean_dem_margin)

# create linear model
rdi_polls_sum_seat_lm <- lm(D_seats ~ mean_dem_margin + rdi_change_pct,
                            data = polls_rdi_election_seat_sum)

summary(rdi_polls_sum_seat_lm)

# Create data frame of current statistics
# dem margin: RCP average = D+0.3
# According to BEA, cpi change from 4/1/22 to 8/1/22 is a 1.91% increase
poll_and_rdi_data_vote <- data.frame(mean_dem_margin = 0.3,
                                rdi_change_pct = 1.91)

# predict Dem seat share with current RDI and Dem Margin 
predict(rdi_polls_sum_seat_lm,
        poll_and_rdi_data_vote,
        interval = "prediction",
        level = 0.95)
```

Below includes how well these models stack up against historical data. What is evident, after including polling data with my economic data, is how these models are not exact, but they are matching the trends much better, compared to last weeks economy-focused model. These models are efficient at predicting if the Democrats will lose the vote share and seat share, only missing the vote share prediction once and the seat share prediction twice. However, the exact vote share percentage and seat share count can be more precise.

#### How My Model Stacks Up Against the Past
```{r, echo = FALSE, message = FALSE}
# True = line
# prediction = dots
ggplot(data = polls_rdi_election_vote_sum,
       mapping = aes(x = year,
                     y = D_majorvote_pct)) +
  geom_line(aes(y = D_majorvote_pct),
            size = 1.5) +
  geom_point(y = predict(rdi_polls_sum_vote_lm, polls_rdi_election_vote_sum),
             color = "red",
             size = 2) +
  geom_hline(yintercept = 50, lty = 2) +
  labs(x = "Year",
       y = "Democratic House Incumbent Party Vote Share ",
       subtitle = "Vote Share Based on RDI Change and Generic Ballot Polling Aggregates",
       title = "True Results (Line) vs Model Prediction (Dots)") +
  theme_bw() +
  theme(axis.text = element_text(size = 10))


# visualize rdi data for Dems seat share under Dem House
ggplot(data = polls_rdi_election_seat_sum,
       mapping = aes(x = year,
                     y = D_seats)) +
  geom_line(aes(y = D_seats),
            size = 1.5) +
  geom_point(y = predict(rdi_polls_sum_seat_lm, polls_rdi_election_seat_sum),
             color = "red",
             size = 2) +
  geom_hline(yintercept = 218, lty = 2) +
  labs(x = "Year",
       y = "Democratic House Incumbent Party Seat Share ",
       subtitle = "Seat Share Based on RDI Change and Generic Ballot Polling Aggregates",
       title = "True Results (Line) vs Model Prediction (Dots)") +
  theme_bw() +
  theme(axis.text = element_text(size = 10))
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# Residuals for Vote Share
ggplot(data = rdi_polls_sum_vote_lm,
       mapping = aes(x = rdi_polls_sum_vote_lm$model$D_majorvote_pct -
rdi_polls_sum_vote_lm$fitted.values)) +
  geom_histogram(bins = 15) +
  labs(x = "Residuals Between True and Predicted Vote Shares",
       y = "Frequency",
       subtitle = "Vote Share Based on RDI Change and Polling Averages",
       title = "Error in Model") +
  theme_bw() +
  theme(axis.text = element_text(size = 10))

# Residuals for Seat Share
ggplot(data = rdi_polls_sum_seat_lm,
       mapping = aes(x = rdi_polls_sum_seat_lm$model$D_seats -
rdi_polls_sum_seat_lm$fitted.values)) +
  geom_histogram(bins = 15) +
  labs(x = "Residuals Between True and Predicted Seat Shares",
       y = "Frequency",
       subtitle = "Seat Share Based on Real Disposable Income Change",
       title = "Error in Model") +
  theme_bw() +
  theme(axis.text = element_text(size = 10))
```

Using these models that historically fit the data well, I will update my forecast with the predictions shown below.

## Updated Forecast:


#### Predicted Democratic Party Vote Share
###### Based on Real Disposable Income and Polling Averages
```{r, echo = FALSE, message = FALSE}
# predict Dem vote share with current RDI and Dem Margin
predict(rdi_polls_sum_vote_lm,
        poll_and_rdi_data_vote,
        interval = "prediction",
        level = 0.95)
```

#### Predicted Democratic Party Seat Share
###### Based on Real Disposable Income and Polling Averages
```{r, echo = FALSE, message = FALSE}
# predict House incumbent seat share with current RDI 
predict(rdi_polls_sum_seat_lm,
        poll_and_rdi_data_vote,
        interval = "prediction",
        level = 0.95)
```

## Conclusion 
Using [RealClearPolitics' Current Aggregated Generic Congressional Ballot,](https://www.realclearpolitics.com/epolls/other/2022-generic-congressional-vote-7361.html) where Democrats hold an advantage +0.3 points and the current RDI change since April of +1.91%, I predict the Democrats to win a national vote share of 49.29% and 211 seats in the House. If this holds, the Democrats will be demoted to the minority party of the House.

However, something special to note is that my prediction intervals remain wide with only these two variables. Thus, more analysis and refining is needed to narrow this interval, since the upper limit of my interval includes the Democrats winning an overall 244 seats (which is still vastly unrealistic and a pipedream for the party, at the moment). 


##### Notes:

*This blog is part of a series of articles meant to progressively understand how election data can be used to predict future outcomes. I will add to this site on a weekly basis under the direction of [Professor Ryan D. Enos](https://www.ryandenos.com). In the run-up to the midterm elections on November 8, 2022, I will draw on all I have learned regarding what best forecasts election results, and I will predict the outcome of the NE-02 U.S. Congressional Election.*

[My project repository can be viewed here.](https://github.com/luke-tomes/election-blog)

##### Sources:

[Bureau of Economic Analysis, RDI Change](https://www.bea.gov/news/2022/personal-income-and-outlays-july-2022)