---
title: 'Blog Post #6 - Ground Game'
author: Ethan Kelly
date: '2022-10-19'
slug: blog-post-6-ground-game
categories: []
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2022-10-19T12:55:26-04:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
## Introduction

  Throughout the 21st Century, turnout across the United States has fluctuated significantly. For the sake of this blog post, voter turnout is described as the % of the amount of voters vs. the voting eligible population (18+ & have the ability to vote). Taking data from the University of California Santa Barbara’s American Presidency Project, which collects data pertaining to presidential elections, we find that in the 2000 election, voter turnout was just 55.3%. Based on ElectProject.org, in the 2020 election, voter turnout was 66.6%; the highest since the 1900 presidential election.

	However, our focus is on the 2022 midterm elections & understanding the impact of turnout on congressional races. For the sake of context, turnout in 2018 was 50.0%, contrary to 2014’s turnout at 36.7%. Both of these races have something in common: they are midterm elections. Midterm elections rarely rival presidential election turnout — but when they do, elections become significantly more interesting.

	The old rule of thumb was that increasing turnout meant an increase in Democratic vote share. Many pundits have been under the assumption that democrats rely heavily on turnout in order to win their elections, whereas republicans are the baseline winners. Taking 2014 & 2018 for instance — 2014 was an overwhelming red wave year, where Democrats were far from enthused & didn’t turnout. 2018, on the other hand, was a significantly high turnout year, in which Democrats won back control of the House of Representatives. Though, it is not that simple.

	In this blog, we will be exploring the impact of turnout on the midterm elections, though we will be considering more than just that. In addition to turnout, we will be mapping our data based on expert predictions and incumbency. In this blog post, we will be utilizing 2018 as our data point — given that we are only using data for all 435 districts for 2018. For incumbency, we will be treating that as a factor by congressional district, denoting incumbency by the incumbent party in the district (even if the race is an open seat). Finally, our expert predictions will be taken from the Cook Political Report, Larry Sabato’s Crystal Ball, and Inside Elections. 

## Background on 2018 Midterm Elections ## 

	To start, here are 2 maps of the 2018 midterm election results, as a baseline to see the election results. The first will show the margin of Democratic Party vote share, whereas the other will show the election winner in red or blue.
	
```{r}
# Loading in the data 
knitr::opts_chunk$set(echo = FALSE)

# Loading in necessary libraries
library(tidyverse)
library(janitor)
library(ggplot2)
library(sf)

# Reading in the data
expert_ratings <- read_csv("expert_rating.csv")
historical_results <- read_csv("house party vote share by district 1948-2020.csv") %>% 
  clean_names()
ratings_2018 <- read_csv("ratings_2018_share_fix.csv")
ratings_2018_fix <- read_csv("ratings_2018_share_fix_1.csv")
```

```{r}
require(tidyverse)
require(ggplot2)
require(sf)
# load geographic data
get_congress_map <- function(cong=114) {
  tmp_file <- tempfile()
  tmp_dir  <- tempdir()
  zp <- sprintf("https://cdmaps.polisci.ucla.edu/shp/districts114.zip",cong)
  download.file(zp, tmp_file)
  unzip(zipfile = tmp_file, exdir = tmp_dir)
  fpath <- paste(tmp_dir, sprintf("districtShapes/districts114.shp",cong), sep = "/")
  st_read(fpath)
}

# load 114th congress
cd114 <- get_congress_map(114)

# vote data
house_party_vote_share_by_district_1948_2020 <- read_csv("house party vote share by district 1948-2020.csv")
h <- house_party_vote_share_by_district_1948_2020
D_2018 <- h %>%
  filter(raceYear == 2018) %>%
  select(raceYear, State, district_num, district_id, RepVotesMajorPercent, DemVotesMajorPercent) %>%
  # summarize party vote share by district
  group_by(district_num, State, district_id) %>%
  summarise(Dem_votes_pct = DemVotesMajorPercent) %>%
  # rename district variable name to match shapefile
  rename(DISTRICT = district_num, STATENAME = State)

# merge
cd114$DISTRICT <- as.numeric(cd114$DISTRICT)
cd114 <- cd114 %>% left_join(D_2018, by=c("DISTRICT", "STATENAME"))
head(cd114$Dem_votes_pct)

# Plotting 2018 Map Based on Results
districts_simp <- rmapshaper::ms_simplify(cd114, keep = 0.01)

ggplot() + 
  geom_sf(data=districts_simp,aes(fill=Dem_votes_pct),
          inherit.aes=FALSE,alpha=0.9) + 
  scale_fill_gradient(low = "red", high = "darkblue", limits=c(0,100)) +
  coord_sf(xlim = c(-124.43, -66.57), ylim = c(23, 49), expand = FALSE) +  
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```
```{r}
D_2018_Absolute <- D_2018 %>%
  mutate(value_cut = cut(Dem_votes_pct, breaks = c(0.00,50,100)))
  
# merge
cd114$DISTRICT <- as.numeric(cd114$DISTRICT)
cd115 <- cd114 %>% left_join(D_2018_Absolute, by=c("DISTRICT", "STATENAME"))
head(cd115$value_cut)

# Plotting 2018 Map Based on Results
districts_simpabsolute <- rmapshaper::ms_simplify(cd115, keep = 0.01)

ggplot() -> gg
gg <- gg + geom_sf(data=districts_simpabsolute,aes(fill=value_cut),
          inherit.aes=FALSE,alpha=0.9)
cols <- c("red", "blue")
gg <- gg  + scale_fill_manual(values=cols) + 
  coord_sf(xlim = c(-124.43, -66.57), ylim = c(23, 49), expand = FALSE) +  
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
print(gg)
```
Note on the map above: some of the districts are grayed out (only 3), this is due to a data error not being captured by the R code mapping the districts. These are all Republican victories -- please view them as such.

Overall, Democrats captured 235 House seats, to the Republican Party's 199. This was the most recent midterm election result.

## Model on expert prediction

Now, we move onto expert predictions for 2018 & see what worked versus did not. Here is a map of the 2018 predictions on the House level.

```{r}
D_2018_1 <- h %>%
  filter(raceYear == 2018) %>%
  select(raceYear, State, district_num, district_id, RepVotesMajorPercent, DemVotesMajorPercent) %>%
  # summarize party vote share by district
  group_by(district_num, State, district_id) %>%
  summarise(Dem_votes_pct = DemVotesMajorPercent) %>%
  # rename district variable name to match shapefile
  rename(DISTRICT = district_num, STATENAME = State)

D_2018_1$district_id[grep(pattern = "00", x = D_2018_1$district_id)] <- c("AK01","DE01","MT01","ND01","SD01", "VT01","WY01")
```

```{r}
# merge
expertratings <- D_2018_1 %>% left_join(ratings_2018_fix, by="district_id") 

expertratings1 <- expertratings %>%
  mutate(value_cut1 = cut(avg, breaks = c(0,3.5,7)))

cd116 <- cd114 %>% left_join(expertratings1, by=c("DISTRICT", "STATENAME"))
head(cd116$avg)

# Plotting 2018 Map Based on Expert
districts_simpabsolute1 <- rmapshaper::ms_simplify(cd116, keep = 0.01)

ggplot() -> gg1
gg1 <- gg1 + geom_sf(data=districts_simpabsolute1,aes(fill=value_cut1),
          inherit.aes=FALSE,alpha=0.9)
cols <- c("blue", "red")
gg1 <- gg1  + scale_fill_manual(values=cols) + 
  coord_sf(xlim = c(-124.43, -66.57), ylim = c(23, 49), expand = FALSE) +  
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
print(gg1)
```

```{r}
require(gridExtra)
plot1 <- gg
plot2 <- gg1
grid.arrange(plot1, plot2, ncol=2)
```

## Model including incumbency + expert predictions

## Model including incumbency + expert predictions + turnout

## Limitations

## Conclusions

